---
title: "Casa Alitas Origins, Languages & Destinations Report"
author: "Kendra Lyons"
date: "5/6/2022"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
library(ggrepel)
library(ggthemes)
library(knitr)
library(scales)
library(tidyverse)
library(usmap)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Data Dates: 1/26/21 to 1/20/22

Data for this project were collected by volunteers and staff at Casa Alitas Welcome Center in Tucson, Arizona. 

During the dates these data were collected, approximately 24,000 individuals from at least 37 different countries passed through the Welcome Center. They spoke at least 41 different Languages, and after being dropped off in Tucson or nearby towns by border authorities, they continued on to 47 states and the District of Columbia. 

```{r data, include = FALSE}

individuals <- read_csv("data/individual_all_clean_1_20_22.csv", na = c("", "NA")) %>% 
  rename("GroupID" = ...1) %>%
  select(arrival_date, GroupID, language, country, state_name, state_abb)

countries <- drop_na(read_csv("data/country_counts_1.20.csv")) %>%
    filter(country != "United States")

languages <- drop_na(read_csv("data/language_counts_1.20.csv"))

destinations <- drop_na(read_csv("data/destination_counts_1.20.csv")) %>%
  filter(n != 2) %>%
  mutate(n = case_when(state_name == "Florida" ~ n + 1, # one group of two split between FL and TX
                       state_name == "Texas" ~ n + 1,
                       TRUE ~ n))

# Add values of 0 for states with no travelers
destinations <- destinations %>%
  add_row(state_name = "Hawaii", n = 0, percent = 0.0) %>%
  add_row(state_name = "North Dakota", n = 0, percent = 0.0) %>%
  add_row(state_name = "Wyoming", n = 0, percent = 0.0) %>%
  rename("state" = state_name)

# get state centroids for labels
centroids <- read_csv("data/us_states_centroids.csv") %>%
  rename("state" = full)

# add centroid data to destination data 
map_destinations <- centroids %>%
  left_join(destinations)

# get immigration policy event data 
policy_events <- read_csv("data/selected_policy_events_2021.csv", skip = 2) %>%
  mutate(date = parse_date(date, format = "%m/%d/%Y"),
         category = category)

# select events of interest
key_policy_events <- policy_events %>%
  filter(category %in% c('MPP ("Remain in Mexico")', 
                         'Title 42', 
                         'Temporary Protected Status')) #, 'Detention'

```


## Languages


```{r languages, echo=FALSE}
# set plot theme
theme_set(theme_minimal(14))


# plot arrivals of other language speakers over time
some_langs = languages$language[languages$n > 4]

individuals %>%
  filter(language %in% some_langs) %>%
  ggplot(aes(x = arrival_date, 
             y = factor(language, rev(some_langs)))) +
  geom_point(#aes(color = factor(language, rev(some_langs))), 
             alpha = .2,
             show.legend = FALSE) +
  #scale_color_viridis_d(option = "viridis", begin = 0, end = 1) +
  #scale_color_viridis_d(option = "plasma") +
  theme(plot.background = element_rect(fill = "#f4eee0")) +
  scale_x_date(date_breaks = "1 month", 
               date_labels = "%b",
               expand = c(.02,-.02)) +
  labs(title = "Arrivals over time by language spoken",
       x = element_blank(), 
       y = element_blank())
  
```

The language most commonly spoken by guests at Casa Alitas was Spanish, spoken by nearly 71% of arrivals in 2021. Brazilian Portuguese was the next most common, spoken by just under 20% of guests. Haitian Creole was spoken by around four percent, and Quichua was spoken by one percent of guests. Hindi and Romanian were the only other languages spoken by more than 100 arrivals, but smaller numbers of guests spoke many other languages including a wide variety of indigenous languages. See Table 1 for a complete list.


```{r, lang_table, echo=FALSE}

languages %>%
  kable(caption = "Table 1: Number and Percentage of CAWC Arrivals by Language Spoken")

```


## Countries of Origin

```{r orig_fw, echo=FALSE}
 # filter out countries with few arrivals
top_ctrys <-  countries$country[0:12]
# Top 12 countries of origin FW 
individuals %>%
  filter(country %in% top_ctrys) %>%
  ggplot(aes(x = arrival_date,
             #color = factor(country, rev(top_ctrys))
             ), fill = "black") +
  geom_bar(show.legend = FALSE) + 
  facet_wrap(~factor(country, top_ctrys), 
             scales = "free_y")  +
  #scale_color_viridis_d(option = "viridis", begin = 1, end = 0) +
  theme(plot.background = element_rect(fill = "#f4eee0")) +
  scale_x_date(date_breaks = "3 month", 
               date_labels = "%b",
               expand = c(.02,-.02)) +
  labs(title = "Daily arrivals over time, top origin countries", 
       x = element_blank(), 
       y = "Number of Individuals") 
```

Venezuelans represented the largest percentage of 2021 arrivals at 22.7%, followed by Brazilians at around 20%. Cubans made up almost 14%, Ecuadorians nearly 12%, Mexicans and Nicaraguans around 7%, and Haitians less than 5%. Colombians, Guatemalans, Peruvians, Hondurans and Indians each made up between one and three percent of arrivals. For a complete list of origin countries, see Table 2.


```{r, country_table, echo=FALSE}
countries %>% 
  filter(n > 1) %>%
  kable(caption = "Table 2: Number and Percentage of CAWC Arrivals by Origin Country")
```

### Policy Effects

Policy changes played an key role in determining which migrants were allowed to enter the country at different points in time throughout the year. Three policies had especially obvious impacts. One is the **Migrant Protection Protocols** (MPP), also known as "Remain in Mexico", which forces migrants to await U.S. immigration trials in Mexico, often under exceedingly dangerous conditions. Another important policy is **Title 42** is used to expel migrants, including many asylum-seekers, upon their arrival at the border without the opportunity to request asylum. Migrants from certain countries were eligible for **Temporary Protected Status** due to war, natural disasters or other extreme circumstances in their home countries. Countries eligible for TPS in 2021 include Venezuela, Honduras, Nicaragua, Haiti, El Salvador and Nepal.

```{r orig_policy, echo=FALSE}
# get countries with arrivals greater than 100
some_ctrys <- countries$country[countries$n > 5]
# Arrivals by country over time dot plot
individuals %>%
  filter(country %in% some_ctrys) %>%
  ggplot(aes(x = arrival_date, 
             y = factor(country, rev(countries$country[1:23])))) +
  geom_vline(data = key_policy_events, 
             aes(xintercept = date, 
                 color = category,
                 linetype = category),
             na.rm = TRUE,
             size = 1) +
  #scale_color_manual(values = c("purple", "goldenrod", "darkgreen")) +
  scale_color_manual(values = c( "darkred", "gray40", "black")) + 
  geom_point(alpha = .2,
             show.legend = FALSE) +
  theme(legend.position = "top", 
        legend.justification = 0,
        plot.background = element_rect(fill = "#f4eee0")) +
  labs(title = "Policy change effects by origin country",
       x = element_blank(), 
       y = element_blank(),
       color = element_blank(),
       linetype = element_blank()) +
  scale_x_date(date_breaks = "1 month", 
               date_labels = "%b",
               expand = c(.02,-.02))
```

```{r policy_table, echo=FALSE}
# print table with policy change details
key_policy_events %>%
  kable(caption = "Table 3: Immigration Policy Changes in 2021 (Source: Investigative Reporting Workshop)")

```


###  Origins by Destination States

There were some strong relationships between certain origin countries and destination states. The most common destination for Venezuelans, Cubans and Nicaraguans was Florida, followed by Texas.  Haitians also  primarily went to Florida, but were more likely to go to New York or New Jersey than to Texas. Some Brazilians went to Florida, but they were much more likely to travel to the East coast, especially to Massachusetts, but also to Pennsylvania, New Jersey and, to a lesser extent, Connecticut. Ecuadorians and Peruvians were also usually destined the East coast, mainly New York and New Jersey. Mexicans most commonly went to California, followed by Texas. Colombians were about equally likely to go to New York, Florida or California. Guatemalans and Indians both commonly went to California, and Hondurans were most likely to go to Texas. 

```{r destorig, echo=FALSE}
# get top destination states
top_states <- destinations$state[1:12]

# plot comparison of top origin countries and destination states
individuals %>%
  filter(country %in% top_ctrys,
         state_name %in% top_states) %>%
ggplot(aes(y = factor(country, rev(top_ctrys)),
           fill = factor(state_abb, levels = rev(destinations$state_abb[1:12])))) +
  geom_bar(position = "fill", na.rm = TRUE) + # normalize bar height
  theme(axis.text.x=element_blank(),
        legend.text = element_text(size=10),
        legend.position = "top",
        legend.justification = 0,
        legend.spacing.x = unit(.5, 'cm'),
        plot.background = element_rect(fill = "#f4eee0")) +
    guides(fill = guide_legend(nrow = 1, reverse = TRUE, 
                             label.position = "top")) +
  scale_fill_grey(start = 1, end = 0) +
  #scale_fill_viridis_d(option = "plasma") +
  labs(title = "Top destination states by origin country", 
       y = element_blank(), 
       x = element_blank(), 
       fill = element_blank())

```

## Destination States

By far, the most popular destination of Casa Alitas guests was Florida, with about 27% travelling there. Massachusetts and New York were the next most popular, with 9.7% and 9.4%, respectively. Texas had 8.5%, New Jersey 6.3% and California 5.7%. Three percent of guests went to Pennsylvania, 2.6% to Georgia, and 2.3% to both Connecticut and Illinois. No other states received more than 2% of arrivals, and only three states had zero arrivals (Hawaii, North Dakota and Wyoming). See Table 4 for a list of numbers and percentages of destinations for all 2021 arrivals 

```{r destmap, echo=FALSE}
# plot data on U.S. map with descriptive labels
plot_usmap(data = destinations, 
           values = "n") +
  theme(legend.position = "right", 
        title = element_text(size = 14)) +
<<<<<<< HEAD
  theme(plot.background = element_rect(fill = "#f4eee0", color = "#f4eee0"),
        legend.background = element_rect(fill = "#f4eee0")) +
  #scale_fill_viridis_c(option = "viridis", begin = 1, end = 0) +
  scale_fill_gradient(low="#f4eee0", high='grey10') +
=======
  scale_fill_viridis_c(option = "plasma") +
>>>>>>> 2a0cbc40c0ad6dddf8d276b980e9caf5f2780c60
  geom_label_repel(data = map_destinations %>% filter(n != 2), 
                   max.overlaps = 21,
                   fontface = "bold",
                   size = 2.8,
                   aes(x = x,
                       y = y,
                       label = paste(abbr, comma(n, accuracy=1)))) +
  labs(title = "Destination states of Casa Alitas guests",
       fill = "People") # MAKE LARGER!
  # scale_fill_continuous(name = "Guests",
  #                        low = "white",
  #                        high = "darkgreen")

```

Logically, the most people went to the most populated states, where they were most likely to know someone willing to be their legal sponsor in the U.S. for the duration of the asylum process. Things didn't always go smoothly with sponsors, though. Sometimes they backed out, leaving guests without a place to go. When that happened, people at Alitas tried to locate another sponsor for them. 


```{r desttl, echo=FALSE}
# vector with top (12) destination states
popular_dests <- destinations$state[destinations$n>100]

individuals %>% 
  filter(state_name %in% popular_dests) %>%
  ggplot(aes(x = arrival_date, 
             y = factor(state_name, rev(popular_dests)))) +
  geom_point(#aes(color = factor(state_name, rev(popular_dests))), 
             alpha = .2, na.rm = TRUE,
             show.legend = FALSE) +
<<<<<<< HEAD
  #scale_color_viridis_d(option = "plasma") +
  theme(plot.background = element_rect(fill = "#f4eee0")) +
=======
  scale_color_viridis_d(option = "plasma") +
>>>>>>> 2a0cbc40c0ad6dddf8d276b980e9caf5f2780c60
  scale_x_date(date_breaks = "1 month", 
               date_labels = "%b",
               expand = c(.02,-.02)) +
  labs(title = "Arrivals over time by destination state",
       x = element_blank(), 
       y = element_blank())
```

At least one guest traveled to every single state except Hawaii, North Dakota and Wyoming. Twenty-eight states received more than 100 migrants who passed through Casa Alitas during their journey. 

```{r, destTable, echo=FALSE}
destinations %>%
  filter(n > 0) %>%
  kable(caption = "Table 4: Number and Percentage of CAWC Arrivals by Destination State")
```

## To Learn More

- **Arizona Public Media Video Feature**: https://www.youtube.com/watch?v=6FUyY0_exoo

- **Casa Alitas Program Website**: https://www.casaalitas.org/

- **Catholic Community Services of Southern Arizona**: https://www.ccs-soaz.org/agencies-ministries/detail/alitas-aid-for-migrant-women-and-children


### Asylum

- **Southern Border Communities Coalition**: https://www.southernborder.org/border_lens_asylum_seekers_at_the_border

- **U.S. Citizenship and Immigration Services**: https://www.uscis.gov/humanitarian/refugees-and-asylum/asylum/obtaining-asylum-in-the-united-states

- **Syracuse University**: https://trac.syr.edu/immigration/reports/667/


### Migration Policy Changes

- **Boundless**: https://www.boundless.com/blog/biden-immigration-tracker/

- **Center for Migration Studies**: https://cmsny.org/biden-immigration-executive-actions/

- **Immigration Policy Tracking Project**: https://immpolicytracking.org/home/

- **Investigative Reporting Workshop**: https://investigativereportingworkshop.org/news/timeline-bidens-immigration-policy/

- **Lawfare Blog**: https://www.lawfareblog.com/president-bidens-immigration-executive-actions-recap#Asylum

- **Pew Research**: https://www.pewresearch.org/fact-tank/2022/01/11/key-facts-about-u-s-immigration-policies-and-bidens-proposed-changes/


### "Remain in Mexico" (MPP)

- **Department of Homeland Security**: https://www.dhs.gov/migrant-protection-protocols

- **Migration Policy Institute**: https://www.migrationpolicy.org/article/court-order-relaunch-remain-in-mexico

- **Reuters**: https://www.reuters.com/world/us/biden-administration-issues-new-memo-ending-trump-remain-mexico-policy-2021-10-29/


### Temporary Protected Status

- **Federal Register**: https://www.federalregister.gov/documents/2021/09/10/2021-19617/continuation-of-documentation-for-beneficiaries-of-temporary-protected-status-designations-for-el

- **National Immigration Forum**: https://immigrationforum.org/article/fact-sheet-temporary-protected-status/

- **U.S. Citizenship and Immigration Services**: https://www.uscis.gov/humanitarian/temporary-protected-status


### Title 42

- **American Immigration Council**: https://www.americanimmigrationcouncil.org/research/guide-title-42-expulsions-border

- **CBS News**: https://www.cbsnews.com/news/immigration-title-42-senate-bill-border-expulsions/

- **Deseret News**: https://www.deseret.com/2022/4/8/23015106/what-is-title-42-and-why-does-biden-plan-to-end-it-your-questions-answered-us-mexico-border

- **El Paso Matters**: https://elpasomatters.org/2021/10/08/timeline-how-title-42-came-to-be-used-on-asylum-seekers/

- **Puente Arizona**: https://puenteaz.org/title42/

- **Washington Office on Latin America**: https://www.wola.org/analysis/a-tragic-milestone-20000th-migrant-deported-to-haiti-since-biden-inauguration/


### News 

- **Boundless**: https://www.boundless.com/blog/boundless-weekly-immigration-news/

- **FEMA Funds for Migrant Aid**: https://tucson.com/news/local/pima-county-uncertain-federal-funds-to-support-asylum-seekers-will-continue/article_1d07a4c8-901d-11ec-8e6c-a37e4491dcf4.html

- **Human Rights Watch**: https://www.hrw.org/news/2021/03/05/mexico-abuses-against-asylum-seekers-us-border

- **International Rescue Committee**: https://www.rescue.org/article/what-happening-us-southern-border
  - En espanol: https://www.rescue.org/article/que-esta-pasando-en-la-frontera-sur-de-estados-unidos

- **Tucson Sentinel**: https://www.tucsonsentinel.com/local/report/030122_casa_alitas_budget/pima-county-spending-13-million-per-month-shelter-migrants-released-by-dhs/

- **Tucson Weekly**: https://www.tucsonweekly.com/tucson/crowded-house-casa-alitas-faces-increased-pressure-as-more-migrants-seek-refugee-status/Content?oid=30449369

